---
title: "Final Animal Abundance"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

# Notebook MT5751 - Project 3 Occupancy Estimation

## I) Notebook Setup

```{r}
# Load libraries
library(tidyverse)
library(corrplot)
library(ggplot2)
library(MuMIn)
library(unmarked)
library(tidyr)
library(car)
library(AICcmodavg)
library(dplyr)

# Load Data
library(statsecol)

# Save the data frame
data(willow)
willow_df <- willow
str(willow_df)

willow_df <- as.data.frame(sapply(willow_df, as.numeric))
```

## II) Data exploration

### A) Distribution of forest Cover

```{r}
ggplot(data = willow_df, aes(x = forest)) +
  geom_histogram(bins = 28, fill = 'lightgreen', color = 'lightgreen') +
  labs(title = "Distribution of Forest Cover", x = "Forest Cover (P of quadrat covered in forest)", y = "Count")
```

### B) Distribution of elevation

```{r}
ggplot(data = willow_df, aes(x = elev)) +
  geom_histogram(bins = 28, fill = 'lightblue', color = 'lightblue') +
  labs(title = "Distribution of Elevation (scaled)", x = "Elevation (scaled)", y = "Count")
```

### D) Distribution of lengh

```{r}
ggplot(data = willow_df, aes(x = elev)) +
geom_histogram(bins = 30,  fill = 'lightgrey', color = 'lightgrey') +
labs(title = "Distribution of lengh (km)", x = "lengh (km)", y = "Count")# Introduction
```

### E) Distribution of intensity per visit

```{r}
# Convert intensity to numeric
willow_df$intensity2 <- as.numeric(willow_df$intensity2)
willow_df$intensity3 <- as.numeric(willow_df$intensity3)

# Reshape data 
willow_df_long <- pivot_longer(willow_df, cols = starts_with("intensity"), names_to = "variable", values_to = "value")

# Create plot
effort_willow_df_long <- ggplot(willow_df_long, aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5) +
  xlab("Intensity") +
  ylab("Density") +
  ggtitle("Density of Intensity Variables") +
  scale_fill_manual(name = "Intensity", values = c("intensity1" = "blue", "intensity2" = "green", "intensity3" = "grey"))

effort_willow_df_long
```

### E) Distribution of Duration per visit

```{r}
# Convert dur to numeric
willow_df$dur2 <- as.numeric(willow_df$dur2)
willow_df$dur3 <- as.numeric(willow_df$dur3)

# Reshape data 
willow_df_dur <- pivot_longer(willow_df, cols = starts_with("dur"), 
                            names_to = "variable", values_to = "value")

# Create a plot
willow_plot_dur <- ggplot(willow_df_dur, aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5) +
  xlab("Value") +
  ylab("Density") +
  ggtitle("Density of Duration Variables") +
  scale_fill_manual(name = "Variable", 
                    values = c("dur1" = "blue", "dur2" = "green", "dur3" = "grey"))

willow_plot_dur
```

### F) Distribution of days since the start of the year on each occasion 

```{r}
# Convert day to numeric
willow_df$day3 <- as.numeric(willow_df$day3)
sapply(select(willow_df, starts_with("day")), class)


# Pivot the data 
willow_df %>%
  pivot_longer(cols = starts_with("day"), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5) +
  labs(x = "Value", y = "Density", title = "Distribution of Days Since the Start of the Year on Each Occasion") +
  scale_fill_manual(name = "Variable", values = c("lightblue", "lightgreen", "lightgrey"))

```

### H) Sampling period

```{r}
# Reshape Data
data_range <- melt(willow_df, id.vars = c("day1", "day2", "day3"),
                  measure.vars = c("y.1", "y.2", "y.3"),
                  variable.name = "survey", value.name = "detection")

# Create a new column
data_range$day_of_survey <- ifelse(data_range$survey == "y.1", data$day1,
                                  ifelse(data_range$survey == "y.2", data$day2, data$day3))

# Aggregate counts
Counts <- data_range %>%
  group_by(day_of_survey) %>%
  summarise(detection_count = sum(detection, na.rm = TRUE))

# Plot
ggplot(Counts, aes(x = day_of_survey, y = detection_count)) +
  geom_line() +
  labs(title = "Daily Counts of Detections Throughout the Survey",
       x = "Day of the Year", y = "Count of Detections") +
  theme_minimal()

```

## III) "unmarked" frame object

```{r}
# Create Dataframe
UM_willow_df <- unmarkedFrameOccu(
  y = willow_df[, c("y.1","y.2","y.3")], 
  siteCovs = data.frame(forest = willow_df$forest,
                        elev = willow_df$elev,
                        length = willow_df$length,
                        elevsq = willow_df$elevsq), 
  obsCovs = list(intensity = willow_df[, c("intensity1", "intensity2", "intensity3")],
                 dur = willow_df[, c("dur1","dur2","dur3")],
                 day = willow_df[, c("day1","day2","day3")])
)
summary(UM_willow_df)
```

## VI) Full Vs Null Model Fit

```{r}
# Full model with all covariates
willow_f <- occu(formula = ~ day + intensity + dur ~ length + forest + elev + elevsq, data = UM_willow_df)   

# Output 
willow_fu <- summary(willow_f)
```

```{r}
# Null model with no covariates
willow_n <- occu(formula = ~ 1 ~ 1, data = UM_willow_df)

# Output
willow_nu <- summary(willow_n)
```

## VI) Model Selection

```{r}
# Define full model with interactions
full <- occu(formula = ~ day + dur + intensity +  day:dur +  day:intensity + intensity:dur ~ length + elev + elevsq + forest + forest:elev + forest:elevsq, data = UM_willow_df)

# dredge 
Dredge_output_with_interaction <- suppressWarnings(dredge(full, rank = "AIC"))

# Print the first 5 rows of the output
Dredge_output_with_interaction[1:5, ]
```

### B) Define best models

```{r}
# Best model 1
Best_1 <- occu(formula = ~ dur + intensity ~  elev + elevsq + forest + (elev:forest) + (elevsq:forest),  data=UM_willow_df)

#  Best model 2
Best_2 <- occu(formula = ~ dur ~  elev + forest + elevsq  + (elev:forest) + (elevsq:forest), data=UM_willow_df)

#  Best model 3
Best_3 <- occu(formula = ~ day + intensity + dur ~  elev + forest + elevsq  + (elev:forest) + (elevsq:forest), data=UM_willow_df)

#  Best model 4
Best_4 <- occu(formula = ~ intensity + dur ~  length + elev + forest + elevsq  + (elev:forest) + (elevsq:forest), data=UM_willow_df)

#  Best model 5
Best_5 <- occu(formula = ~ intensity + dur ~  elev + forest + elevsq  + (elev:forest) , data=UM_willow_df)

#  Best model 6
Best_6 <- occu(formula = ~ intensity + dur ~  elev + forest + elevsq, data=UM_willow_df)
```

### C) regularized occupancy

```{r}
# Define a range of lambda values to test
lamb_vals <- 10^seq(-4, -1, length.out = 30)

# Initialize vector
AIC_vals <- numeric(length(lamb_vals))

# Loop over lambda values, fit model each time, and record AIC
for (i in seq_along(lamb_vals)) {
    model <- occuPEN(~ intensity + dur ~ elev + forest + elevsq + (elev:forest), 
                     data = UM_willow_df, 
                     lambda = lamb_vals[i])
    AIC_vals[i] <- AIC(model)
}

# Find lambda with lowest AIC
min_lamb <- lamb_vals[which.min(AIC_vals)]

# Fit best model
Best_5_reg <- occuPEN(~ intensity + dur ~ elev + forest + elevsq + (elevsq:forest), 
                      data = UM_willow_df, 
                      lambda = min_lamb)

```

## VI) Model Evaluation

### **A) Multicolinearity**

```{r}
# Best model 1
vif_results <- vif(Best_1, type = 'state')
print("VIF Results for Best Model 1:")
print(vif_results)
vif_results_1 <- vif(mod = Best_1, type = 'det')
print(vif_results_1)

# Best model 2
print("VIF Results for Best Model 2:")
vif_results_1 <- vif(mod = Best_2, type = 'state')
print(vif_results_1)

# Best model 3
vif_results <- vif(mod = Best_3, type = 'state')
print("VIF Results for Best Model 3:")
print(vif_results)
vif_results_1 <- vif(mod = Best_3, type = 'det')
print(vif_results_1)

# Best model 4
vif_results <- vif(mod = Best_4, type = 'state')
print("VIF Results for Best Model 4:")
print(vif_results)
vif_results_1 <- vif(mod = Best_4, type = 'det')
print(vif_results_1)

# Best model 5
vif_results <- vif(mod = Best_5, type = 'state')
print("VIF Results for Best Model 5:")
print(vif_results)
vif_results_1 <- vif(mod = Best_5, type = 'det')
print(vif_results_1)

# Best model 5
vif_results <- vif(mod = Best_5, type = 'state')
print("VIF Results for Best Model 5:")
print(vif_results)
vif_results_1 <- vif(mod = Best_5, type = 'det')
print(vif_results_1)

# Best model 6
vif_results <- vif(mod = Best_6, type = 'state')
print("VIF Results for Best Model 6:")
print(vif_results)
vif_results_1 <- vif(mod = Best_6, type = 'det')
print(vif_results_1)
```

### B) Goodness-of-fit

##### **I) Model with both interactions**

```{r}
# Create a suppress warning wrapper
suppressWarningsWrapper <- function(...) {
  suppressWarnings({
    mb.gof.test(...)})}

GOF_WILLOW_1 <- suppressWarningsWrapper(Best_1, nsim = 5, plot.hist = FALSE)
print(GOF_WILLOW_1)
```

##### **II) Model with one interactions**

```{r}
GOF_WILLOW_2 <- suppressWarningsWrapper(Best_5, nsim = 5, plot.hist = FALSE)

print(GOF_WILLOW_2)
```

##### **III) Model with no interaction**

```{r}
GOF_WILLOW_3 <- suppressWarningsWrapper(Best_5, nsim = 50, plot.hist = FALSE)

print(GOF_WILLOW_3)
```

### B) Cross validation

##### **I) Model with both interactions**

```{r}
(kfold_1 = crossVal(Best_1, method="Kfold", folds=10))
```

##### **II) Model with one interactions**

```{r}
(kfold_2 = crossVal(Best_5, method="Kfold", folds=10))
```

##### **III) Model with no interaction**

```{r}
(kfold_3 = crossVal(Best_6, method="Kfold", folds=10))
```

##### **IV) Model With one interaction and regularization**

```{r}
(kfold_3 = crossVal(Best_5_reg, method="Kfold", folds=10))
```

```{r}
summary(Best_5)
```

## VII) Effects plots

### A) Detection

#### a) Duration

```{r}
# Create a dummy data frame with the mean intensity
Duration_df <- data.frame(dur = seq(
  min(UM_willow_df@obsCovs$dur, na.rm=TRUE),
                                   max(UM_willow_df@obsCovs$dur, na.rm=TRUE),
                                   length = 30),  
                         intensity = mean(UM_willow_df@obsCovs$intensity, na.rm=TRUE))

# Create predictions
Duration_df <- predict(Best_5,type="det",newdata = Duration_df, append = TRUE)  

# Plot the results
Effects_duration <- ggplot(data = Duration_df, aes(x = dur, y = Predicted)) +
  geom_ribbon(aes(ymin=lower, ymax=upper), fill="#808080", alpha=0.1) +
  geom_line(size=1,color="#ADD8E6") +
  ylab("Pr(Detection)") + xlab("Survey Duration (Min)") + ylim(0,1) + theme_bw()

Effects_duration
```

#### b) Intensity

```{r}
# Create dummy data frame
Intensity_df <- data.frame(
  dur = mean(UM_willow_df@obsCovs$dur, na.rm = TRUE),  
  intensity = seq(
    min(UM_willow_df@obsCovs$intensity, na.rm = TRUE),
    max(UM_willow_df@obsCovs$intensity, na.rm = TRUE),
    length.out = 30))

# Create predictions
Intensity_df <- predict(Best_5,type="det",newdata = Intensity_df, append = TRUE)  

# Plot the results
Effects_Intensity <- ggplot(data = Intensity_df, aes(x = intensity, y = Predicted)) +
  geom_ribbon(aes(ymin=lower, ymax=upper), fill="#808080", alpha=0.1) +
  geom_line(size=1,color="#228B22") +
  ylab("Pr(Detection)") + xlab("Intensity of search (duration/length) ") + ylim(0,1) + theme_bw()

Effects_Intensity
```

### A) Occupancy

#### a) Elevation

```{r}
# Set mean and standard deviation of elevation
avgElevation <- 1182.574
elevationSD <- 646.333

# Create sequence so that it can be squared
elev_seq <- seq(min(UM_willow_df@siteCovs$elev, na.rm = TRUE),
                max(UM_willow_df@siteCovs$elev, na.rm = TRUE),
                length = 30)

# Create dummy data frame
elevation_df <- data.frame(
  elev = elev_seq,
  elevsq = elev_seq^2,  # Square elevation
  forest = mean(UM_willow_df@siteCovs$forest, na.rm = TRUE))

# Create predictions
elevation_df <- predict(Best_5, type = "state", newdata = pred_psi_ele, append = TRUE)

# Plot the results
elevation_effects <- ggplot(data = elevation_df, aes(x = (elev * elevationSD + avgElevation),, y = Predicted)) +
  
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#5B84B1FF", alpha = 0.1) +
  geom_line(size = 1, color = "#5B84B1FF") +
  ylab("Pr(Occupancy)") + xlab("Elevation (m)") + ylim(0, 1) + theme_bw()

elevation_effects

```

#### C) Forest

```{r}
# Create prediction data frame
Forest_df <- data.frame(
  forest = seq(min(UM_willow_df@siteCovs$forest, na.rm = TRUE),
                  max(UM_willow_df@siteCovs$forest, na.rm = TRUE),
                  length = 30),
  elev = mean(UM_willow_df@siteCovs$elev, na.rm = TRUE),
  elevsq = (mean(UM_willow_df@siteCovs$elev, na.rm = TRUE))^2)

# Generate predictions
Forest_df <- predict(Best_5, type = "state", newdata = Forest_df, append = TRUE)

# Create plot
Plot_forest <- ggplot(data = Forest_df, aes(x = forest, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#228B22", alpha = 0.1) +
  geom_line(size = 1, color = "#228B22") +
  ylab("Pr(Occupancy)") + xlab("Forest") + ylim(0, 1) + theme_bw()

Plot_forest
```

```{r}
library(gridExtra)

combined_plot <- grid.arrange(Plot_forest, elevation_effects, Effects_Intensity, Effects_duration, ncol = 2, nrow = 2)
ggsave("combined_plot.png", combined_plot, width = 10, height = 10)

```

## VII) Prediction

### A) Detection prediction

```{r}
# Define quadrats 
quadList <- c(25, 62, 150, 203)
# Initialize lists
detectionPred <- vector("list", length = length(quadList))
detectionCounter <- 1

# Loop for each quadrat
for (quad in quadList) {
  # Data values 
  quadData <- data.frame(intensity = UM_willow_df@obsCovs$intensity[quad],
                         dur = UM_willow_df@obsCovs$dur[quad], 
                         elev = UM_willow_df@siteCovs$elev[quad],
                         elevsq = UM_willow_df@siteCovs$elevsq[quad],
                         forest = UM_willow_df@siteCovs$forest[quad])
  
    # Prediction
  detectionPred[[detectionCounter]] <- predict(Best_5, type = "det", newdata = quadData, append = TRUE)
  detectionCounter <- detectionCounter + 1}

# Format df
detectionDF <- do.call(rbind, lapply(detectionPred, unlist)) %>% round(3)
rownames(detectionDF) <- paste0("Quad ", quadList)
colnames(detectionDF) <- c("PredictedVal", "Std", "Lower", "Upper", "Int", "Dur", "Elev", "EleS", "Forest")

# Adjusting elevation 
detectionDF[, "Elev"] <- detectionDF[, "Elev"] * elevationSD + avgElevation

detectionDF
```

### A) Occupancy prediction

```{r}
# Initialize lists 
occupancyPred <- vector("list", length = length(quadList))
occupancyCounter <- 1

# Loop for each quadrat
for (quad in quadList) {
  # apend values to data frame
  quadData <- data.frame(intensity = UM_willow_df@obsCovs$intensity[quad],
                         dur = UM_willow_df@obsCovs$dur[quad],
                         elev = UM_willow_df@siteCovs$elev[quad],
                         elevsq = UM_willow_df@siteCovs$elevsq[quad],
                         forest = UM_willow_df@siteCovs$forest[quad])
  # Prediction
  occupancyPred[[occupancyCounter]] <- predict(Best_5, type = "state", newdata = quadData, append = TRUE)
  occupancyCounter <- occupancyCounter + 1}

# Format df
occupancyDF <- do.call(rbind, lapply(occupancyPred, unlist)) %>% round(3)
rownames(occupancyDF) <- paste0("Quad ", quadList)
colnames(occupancyDF) <- c("PredictedVal", "Std", "Lower", "Upper", "Int", "Dur", "Elev", "EleS", "Forest")


# Adjusting elevation
occupancyDF[, "Elev"] <- occupancyDF[, "Elev"] * elevationSD + avgElevation


occupancyDF
```
